Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS: remove void type, still needs to come back and explain a little better how nil* works

\newcommand{\Value}{\mathbf{value}}
\newcommand{\Any}{\mathbf{any}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\Self}{\mathbf{self}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Integer}{\mathbf{integer}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\mathbf{void}}
\newcommand{\Const}{\mathbf{const}}

@@

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end either with $\Void$ or with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.

@@

Typed Lua preserves this behavior in assignment, but allows
both functions with strict arity (which use $\Void$) and
functions with flexible arity (which use variadic types).

The types of expression lists always end in a variadic type
that is $\Nil{*}$ if the number of values the list produces
is statically known. Rule \textsc{S-VARNIL-VOID} makes lets
an expression list with a statically known number of
values be used as the arguments or return values of a function
with a strict arity:
\[
\begin{array}{c}
\mylabel{S-VARNIL-VOID}\\
\senv \vdash \Nil{*} \subtype \Void
\end{array}
\]

>>>>

\newcommand{\Value}{\mathbf{value}}
\newcommand{\Any}{\mathbf{any}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\Self}{\mathbf{self}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Integer}{\mathbf{integer}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\Nil{*}}
\newcommand{\Const}{\mathbf{const}}

@@

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.

@@

Typed Lua preserves this behavior in assignment, but allows
both functions with strict arity, which use $\Void$, and
functions with flexible arity, which use other variadic types,
and can use $\Value{*}$ if they want to just ignore extra
arguments.

The types of expression lists always end in a variadic type
that is $\Nil{*}$ if the number of values the list produces
is statically known.
